Measuring causal influence with\\ back-to-back regression

### 2019-05-23 paper/introduction.tex
Added citation placeholders.

First, causal and non-causal factors may be numerous and collinear. In physics, for example, a student may have difficulty identifying whether fusion is caused by a change in temperature or in pressure, as these two factors can be difficult to manipulate independently. This issue becomes increasingly pronounced as the number of potential factor increases. In neuroscience, for example, it can be challenging to identify whether the frequency of a word presented on a subject's retina modulates brain activity. Indeed, word frequency strongly covaries with word length (short words are more frequent than long words) as well as with word category (determinants are more frequent than adverbs), the individual letters from which it is composed etc (e.g. words that contain "z" are rare) and so forth. As it can be particularly difficult to generate a set of words that control for all of these factors simultaneously, it is common to use forward modeling, i.e. to train a model to predict observations (e.g. brain activity) from a minimal combination of competing factors (e.g. word length, word frequency, etc), and investigate the resulting parameters of the best model.

The second challenge to measuring causal influence is that observations are becoming increasingly large and complex. The relationship between causes and effects is thus often considered in a backward manner, by training models to maximally predict causes from large and complex observations. For example, brain activity is generally recorded with hundreds of sensors simultaneously. It is thus common to use 'decoding' techniques, by, for example, fitting a support vector machine across multiple sensors to predict the category of a stimulus.
%
Both forward and backward modeling have competing benefits and drawbacks. Specifically, forward modeling disentangles the independent contribution of collinear factors, but does not combine multidimensional observations. By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors. \par

In this paper, we introduce the 'back-to-back regression', which combines the benefits of forward and backward modeling. After detailing the method and proving its convergence, we show with synthetic data that it outperforms state-of-the-art forward, backward and cross-decomposition techniques, both in terms of estimating causal influence and out-of-environment predictions. Finally, we apply back-to-back regression to a large neuroimaging dataset and reveal that distinct but collinear word features leads to remarkably distinguishable brain responses.

>>>>

First, causal and non-causal factors may be numerous and collinear. In physics, for example, a student may have difficulty identifying whether fusion is caused by a change in temperature or in pressure, as these two factors can be difficult to manipulate independently. This issue becomes increasingly pronounced as the number of potential factor increases. In neuroscience, for example, it can be challenging to identify whether the frequency of a word presented on a subject's retina modulates brain activity [CITE]. Indeed, word frequency strongly covaries with word length (short words are more frequent than long words) as well as with word category (determinants are more frequent than adverbs), the individual letters from which it is composed etc (e.g. words that contain "z" are rare) and so forth. As it can be particularly difficult to generate a set of words that control for all of these factors simultaneously, it is common to use forward modeling, i.e. to train a model to predict observations (e.g. brain activity) from a minimal combination of competing factors (e.g. word length, word frequency, etc), and investigate the resulting parameters of the best model [CITE].

The second challenge to measuring causal influence is that observations are becoming increasingly large and complex. The relationship between causes and effects is thus often considered in a backward manner, by training models to maximally predict causes from large and complex observations. For example, brain activity is generally recorded with hundreds of sensors simultaneously. It is thus common to use 'decoding' techniques, by, for example, fitting a support vector machine across multiple sensors to predict the category of a stimulus [CITE].
%
Both forward and backward modeling have competing benefits and drawbacks. Specifically, forward modeling disentangles the independent contribution of collinear factors, but does not combine multidimensional observations. By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors [CITE].

We introduce the 'back-to-back regression', which combines the benefits of forward and backward modeling (Section~\ref{sec:algorithm}) After detailing the method and proving its convergence (Section~\ref{sec:theorem}), we show with synthetic data that it outperforms state-of-the-art forward, backward and cross-decomposition techniques, both in terms of estimating causal influence and out-of-environment predictions (Section~\ref{sec:experiment_synthetic}). Finally, we apply back-to-back regression to a large neuroimaging dataset and reveal that distinct but collinear word features leads to distinguishable brain responses (Section~\ref{sec:experiment_real}).

### 2019-05-23 paper/paper.tex
Added citation placeholders.

@@

\newcommand{\dlp}[1]{{\color{red} (DLP: #1)}}
\newcommand{\mo}[1]{{\color{green} (MO: #1)}}
\newcommand{\jrk}[1]{{\color{blue} (JRK: #1)}}
\newcommand{\fc}[1]{{\color{orange} (FC: #1)}}

@@

\section{Back to back regression}

@@

We consider the measurement of multivariate signal $Y$, generated from a set of variables $X$, via some unknown linear apparatus $F$.
%
Not all the variables in $X$ exert a causal influence on $Y$.
%
By considering a square binary diagonal matrix of \emph{causal influences} $E$, we denote by $XE$ the causal factors of $Y$.
%
In summary, our model can be formalized as:
%
\begin{equation}
    Y = (XE + N)F + N',\label{eq:model}
\end{equation}
%
where $N$ and $N'$ are some unobserved noise variables.
%
While the triplet of variables $X$, $N$, and $N'$ are independent, we allow each of them to have a general covariance matrix.
%
In practice, we observe $n$ samples $(X, Y)$ from the model.
%
This process, along with the sizes of all variables involved, is illustrated in Figure~\ref{fig:b2b}.

Given the model in Equation~\eqref{eq:model}, the goal of Back-to-Back Regression (B2BR) is to estimate the matrix of causal influences $E$.

@@

If using regularized least-squares, arguably the most commonly employed linear regression technique, the solution of B2BR has a closed form solution:
\begin{align}
    \hat G &= (Y^\top Y + \Lambda_Y)^{-1} Y^\top X,\label{eq:solG}\\
    \hat H &=(X^\top X + \Lambda_X)^{-1} X^\top Y \hat G,\label{eq:solH}
\end{align}
%
where $\Lambda_X$ and $\Lambda_Y$ are two diagonal matrices containing regularization parameters, useful to invert the covariance matrices of $X$ and $Y$ if these are ill-conditioned.

@@

Performing two regressions over the same data sample can result in overfitting, as spurious correlations in the data absorbed by the first regression will be leveraged by the second one.
%
To avoid this issue, we split our sample $(X, Y)$ at random into two equally-sized halves $(X_1, Y_1)$ and $(X_2, Y_2)$.
%
Then, the first regression is performed using $(X_1, Y_1)$, and the second regression is performed using $(X_2, Y_2)$.
%
To compensate for the reduction in sample size caused by the split, B2BR is repeated over many random splits, and the final estimate $\hat E$ of the causal influence matrix is the average over the estimates associated to each split.
%
After obtaining $\hat{E}$, we can fit a final regression from $X \hat{E}$ to $Y$ as a final B2BR prediction rule.

@@

\subsection{Theoretical guarantees}
% As discussed in the introduction, our problem can be formulated (up to a redefinition of the features of X) as $Y=f(EX+N)$, with f an unknown function.
% %
% If F is linear we can rewrite it as $Y = F(EX + N)$, with X an (dx, n) matrix of possible causes, E a (dx,dx) binary diagonal matrix selecting active features of X.
% %
% N a (dx,n) homoscedastic noise matrix, F an unknown (possibly full rank) (dy,dx) transformation corresponding to the measuring apparatus, and Y a (dy, n) matrix of measured effects.
% 
% The two steps of back-to-back regression amounts to finding $\hat G=\arg \min_G \left \| X-GY \right \|^2$ and $\hat H =\arg \min_H \left \| \hat GY - HX \right \|^2$ (all matrix norms henceforth are Frobenius norms).

@@

Recall the rank factorisation property of the pseudo-inverse \citep{todo}: if $A=BC$ with $A\in\mathbb{R}^{m\times n}$ of rank $k$, $B\in\mathbb{R}^{m\times k}$ and $C\in\mathbb{R}^{k\times n}$, then $A^\dagger=C^\dagger B^\dagger$ and, since $B$ is full rank, $B^\dagger=(B^\top B)^{-1} B^\top$.
%
Using this property on the minimal norm solution $\hat G$ yields
\begin{equation}
\begin{aligned}
(XEF)^\dagger X&=((XK^\top)(KF))^\dagger X\\
&=(KF)^\dagger ((XK^\top)^\top (XK^\top))^{-1} (XK^\top)^\top X\\
&=(KF)^\dagger (KX^\top XK^\top)^{-1} KX^\top X
\end{aligned}
\end{equation}

@@

In regression analysis, the traditional approach to this problem consists in testing whether the regression coefficients differ from zero.
%
However, this test will not succeed here, since the estimate $\hat{E}$ is obtained from a double regression, and any employed regularization will add a bias to the diagonal of $\hat{E}$.
%
Instead, we treat the binarization of $\hat{E}$ as a clustering problem: separate the elements in the diagonal into a group of ``small values'', and a group of ``large values''.
%
More specifically, we propose to maximize the ratio of inter-group variance and to minimize the intra-group variance, over all possible splits of the diagonal into $p$ largest values and $d_x-p$ smallest values.
%
Letting $m_0$ and $m_1$ be the average values of the two clusters, $p$ and $d_x-p$ their size, and $v$ the total variance of the sample, we select the split maximizing the Sonquist-Morgan \citep{todo} criterion $\frac{p(d_x-p)}{d_x} \frac{(m_1 - m_0)^2}{v}$.
%
To binarize $\hat{E}$, set to one all the diagonal entries belonging to the ``large values'' group in the decided split, and setting to zero the rest of the diagonal entries.

@@

\subsection{Synthetic data}

@@

\subsection{Magnetoencephalography data}

@@

Back-to-back regression is closely related to Canonical Correlation Analysis \citep[CCA]{todo}.
%
While CCA computes the eigenvector decomposition of \eqref{eq:solH}, while we compute its diagonal.
%
If \eqref{eq:solH} is diagonal, as in our model \eqref{eq:model}, the variables in $X$ are the eigenvectors.
%
If all the associated eigenvalues are either $0$ or $1$, then all the variables are either part of the $Y$ subspace, or orthogonal to it, respectively.
%
This is because these eigenvalues correspond to the squared cosine of the angle between the corresponding variable in $X$ and the $Y$ subspace.
%
Thus, back-to-back regression is equivalent to CCA under the constraint that the subspaces spanned by $X$ and $Y$ are perpendicular, $X$ is spanned by vectors of $Y$ and vectors orthogonal to $Y$, and that the parallel and orthogonal components of $X$ are spanned by disjoint sets of features of $X$.

>>>>

% TODO: Talk about cause-effect methods, trace method, half-sibling
% TODO: Cite Roweis component analysis methods paper
% TODO: review abstract
% TODO: review conclusion

@@

@@

\section{Back to back regression}
\label{sec:algorithm}

@@

We consider the measurement of multivariate signal $Y$, generated from a set of variables $X$, via some unknown linear apparatus $F$.
%
Not all the variables in $X$ exert a causal influence on $Y$.
%
By considering a square binary diagonal matrix of \emph{causal influences} $E$, we denote by $XE$ the causal factors of $Y$.
%
In summary, our model can be formalized as:
%
\begin{equation}
    Y = (XE + N)F + N',\label{eq:model}
\end{equation}
%
where $N$ and $N'$ are some unobserved noise variables.
%
While the triplet of variables $X$, $N$, and $N'$ are independent, we allow each of them to have a general covariance matrix.
%
In practice, we observe $n$ samples $(X, Y)$ from the model.
%
This process, along with the sizes of all variables involved, is illustrated in Figure~\ref{fig:b2b}.
%
Given the model in Equation~\eqref{eq:model}, \textbf{the goal} of Back-to-Back Regression (B2BR) is to estimate the matrix of causal influences $E$.

@@

If using regularized least-squares, arguably the most commonly employed linear regression technique [CITE], the solution of B2BR has a closed form solution:
\begin{align}
    \hat G &= (Y^\top Y + \Lambda_Y)^{-1} Y^\top X,\label{eq:solG}\\
    \hat H &=(X^\top X + \Lambda_X)^{-1} X^\top Y \hat G,\label{eq:solH}
\end{align}
%
where $\Lambda_X$ and $\Lambda_Y$ are two diagonal matrices of regularization parameters, useful to invert the covariance matrices of $X$ and $Y$ if these are ill-conditioned.

@@

Performing two regressions over the same data sample can result in overfitting, as spurious correlations in the data absorbed by the first regression will be leveraged by the second one.
%
To avoid this issue, we split our sample $(X, Y)$ at random into two equally-sized halves $(X_1, Y_1)$ and $(X_2, Y_2)$.
%
Then, the first regression is performed using $(X_1, Y_1)$, and the second regression is performed using $(X_2, Y_2)$.
%
To compensate for the reduction in sample size caused by the split, B2BR is repeated over many random splits, and the final estimate $\hat E$ of the causal influence matrix is the average over the estimates associated to each split [CITE BOOTSTRAP].
%
After obtaining $\hat{E}$, we can fit a final regression from $X \hat{E}$ to $Y$ as a final B2BR prediction rule.

@@

\subsection{Theoretical guarantees}
\label{sec:theorem}
% As discussed in the introduction, our problem can be formulated (up to a redefinition of the features of X) as $Y=f(EX+N)$, with f an unknown function.
% %
% If F is linear we can rewrite it as $Y = F(EX + N)$, with X an (dx, n) matrix of possible causes, E a (dx,dx) binary diagonal matrix selecting active features of X.
% %
% N a (dx,n) homoscedastic noise matrix, F an unknown (possibly full rank) (dy,dx) transformation corresponding to the measuring apparatus, and Y a (dy, n) matrix of measured effects.
% 
% The two steps of back-to-back regression amounts to finding $\hat G=\arg \min_G \left \| X-GY \right \|^2$ and $\hat H =\arg \min_H \left \| \hat GY - HX \right \|^2$ (all matrix norms henceforth are Frobenius norms).

@@

Recall the rank factorisation property of the pseudo-inverse [CITE]: if $A=BC$ with $A\in\mathbb{R}^{m\times n}$ of rank $k$, $B\in\mathbb{R}^{m\times k}$ and $C\in\mathbb{R}^{k\times n}$, then $A^\dagger=C^\dagger B^\dagger$ and, since $B$ is full rank, $B^\dagger=(B^\top B)^{-1} B^\top$.
%
Using this property on the minimal norm solution $\hat G$ yields
\begin{equation}
\begin{aligned}
(XEF)^\dagger X&=((XK^\top)(KF))^\dagger X\\
&=(KF)^\dagger ((XK^\top)^\top (XK^\top))^{-1} (XK^\top)^\top X\\
&=(KF)^\dagger (KX^\top XK^\top)^{-1} KX^\top X
\end{aligned}
\end{equation}

@@

In regression analysis, the traditional approach to this problem employs a $t$-test to check whether the regression coefficients differ from zero [CITE].
%
However, this test will not succeed here, since the estimate $\hat{E}$ is obtained from a double regression, and any employed regularization will add a bias to the diagonal of $\hat{E}$.
%
Instead, we treat the binarization of $\hat{E}$ as a clustering problem: separate the elements in the diagonal into a group of ``small values'', and a group of ``large values''.
%
More specifically, we propose to maximize the ratio of inter-group variance and to minimize the intra-group variance, over all possible splits of the diagonal into $p$ largest values and $d_x-p$ smallest values.
%
Letting $m_0$ and $m_1$ be the average values of the two clusters, $p$ and $d_x-p$ their size, and $v$ the total variance of the sample, we select the split maximizing the Sonquist-Morgan [CITE] criterion $\frac{p(d_x-p)}{d_x} \frac{(m_1 - m_0)^2}{v}$.
%
To binarize $\hat{E}$, set to one all the diagonal entries belonging to the ``large values'' group in the decided split, and setting to zero the rest of the diagonal entries.

@@

\subsection{Synthetic data}
\label{sec:experiment_synthetic}

@@

\subsection{Magnetoencephalography data}
\label{sec:experiment_real}

@@

Back-to-back regression is closely related to Canonical Correlation Analysis [CITE].
%
While CCA computes the eigenvector decomposition of \eqref{eq:solH}, while we compute its diagonal.
%
If \eqref{eq:solH} is diagonal, as in our model \eqref{eq:model}, the variables in $X$ are the eigenvectors.
%
If all the associated eigenvalues are either $0$ or $1$, then all the variables are either part of the $Y$ subspace, or orthogonal to it, respectively.
%
This is because these eigenvalues correspond to the squared cosine of the angle between the corresponding variable in $X$ and the $Y$ subspace.
%
Thus, back-to-back regression is equivalent to CCA under the constraint that the subspaces spanned by $X$ and $Y$ are perpendicular, $X$ is spanned by vectors of $Y$ and vectors orthogonal to $Y$, and that the parallel and orthogonal components of $X$ are spanned by disjoint sets of features of $X$.

### 2019-05-23 paper/synthetic.tex
Added citation placeholders.

We compare the performance of B2BR against five competing methods:
%
Ridge regression \citep{}, multitask Lasso \citep{}, Partial Least Squares or PLS \citep[PLS]{}, Canonical Correlation Analysis \citep[CCA]{}, and Reduced Rank Regression \citep[RRR]{}.
%
Each of these methods estimates a matrix of coefficients $\hat{W} \in \mathbb{R}^{d_x \times d_y}$, from which we estimate the diagonal mask $\hat{E}$ using the feature importances $\hat{E}_{i,i} = \| W_{i, :} \|$.
%
For B2BR, we obtain $\hat{E}$ as described in Algorithm~\ref{algorithm:b2br}.
%
We employ $5$-fold cross-validation to select the optimal hyper-parameters of each method.

>>>>

We compare the performance of B2BR against five competing methods:
%
Ridge regression [CITE], multitask Lasso [CITE], Partial Least Squares or PLS [CITE], Canonical Correlation Analysis or CCA [CITE], and Reduced Rank Regression or RRR [CITE].
%
Each of these methods estimates a matrix of coefficients $\hat{W} \in \mathbb{R}^{d_x \times d_y}$, from which we estimate the diagonal mask $\hat{E}$ using the feature importances $\hat{E}_{i,i} = \| W_{i, :} \|$.
%
For B2BR, we obtain $\hat{E}$ as described in Algorithm~\ref{algorithm:b2br}.
%
We employ $5$-fold cross-validation to select the optimal hyper-parameters of each method.
